The Iterative Signature Algorithm for Gene Expression data

### 2009-10-30 eisa/inst/doc/EISA_tutorial.Rnw
Simplify Makefile, reorganize HTML generation from the vignettes.

\newcommand{\Rfunction}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

\SweaveOpts{cache=TRUE}
\SweaveOpts{prefix.string=graphics/plot}

@@

<<set width,echo=FALSE,print=FALSE>>=
options(width=60)
@

@@

The Iterative Signature Algorithm (ISA) is biclustering method. The
input of a biclustering method a matrix and its output is a set of
biclusters that fulfill some criteria; a bicluster is a block of the
potentially reordered input matrix.

Most commonly, this algorithm is used on microarray expression data,
to find gene sets that are coexpressed across a subset of the original
samples. In the original ISA paper the biclusters are called
transription modules (TM), we will often refer them under this name in
the following.

@@

First, we load the required packages and the data to analyze. ISA is
implemented in the \Rpackage{eisa} and \Rpackage{isa2} packages, see
Section{sec:isapackages} for a more elaborated summary about the two
packages. It is enough to load the \Rpackage{eisa} package,
\Rpackage{isa2} and other required packages are loaded automatically.
\Rpackage{eisa} 
<<load the packages>>=
library(eisa)
@

In this tutorial we will use the data in the \Rpackage{ALL}
package. 
<<load the data>>=
library(ALL)
library(hgu95av2.db)
library(affy)
data(ALL)
@
This is a data set from a clinical trial in acute
lymphoblastic leukemia and it contains \Sexpr{ncol(ALL)} samples
altogether.

@@

The simplest way to run ISA is to choose the two threshold parameters
and then call the \Rfunction{isa} function on the
\Rclass{ExpressionSet} object:
<<simple ISA>>=
thr.gene <- 2.7
thr.cond <- 1.4
set.seed(1) # to get the same results, always
modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
@

This first applies a non-specific filter to the data set filters the
data set and then runs ISA from
\Sexpr{formals(eisa:::ISA)$no.seeds} % $
random seeds (the default). See Section~\ref{sec:detailed-isa} if the
default parameters are not appropriate for you and need more control.

@@

Calling \Rfunction{length} on \variable{modules} returns the number of
ISA modules in the set, \Rfunction{dim} gives the dimension of the
input expression matrix: the number of features and the number of
samples.
<<accessors 1>>=
length(modules)
dim(modules)
@

Functions \Rfunction{featureNames} and \Rfunction{sampleNames} return
the names of the features and samples, just like the functions with
the same name for an \Rclass{ExpressionSet}
<<accessors 2>>=
featureNames(modules)[1:5]
sampleNames(modules)[1:5]
@

The \Rfunction{getNoFeatures} function returns a numeric vector, the
number of features (probesets in our case) in each module. Similarly,
\Rfunction{getNoSamples} returns a numeric vector, the number of
samples in each module. Similarly, \Rfunction{pData} returns the
phenotype data of the expression set as a data frame. The
\Rfunction{getOrganism} function returns the scientific name of the
organism, \Rfunction{annotation} the name of the chip. For the former
the appropriate annotation must be installed.
<<number of features and samples>>=
getNoFeatures(modules)
getNoSamples(modules)
colnames(pData(modules))
getOrganism(modules)
annotation(modules)
@

@@

\Rfunction{getFeatureNames} lists the probes (more precisely,
the feature names coming from the \Rclass{ExpressionSet} object) in
the modules. It returns a list, here we just print the first entry in
the list.
<<list genes in modules>>=
getFeatureNames(modules)[[1]]
@

@@

Remember, that the ISA biclustering is not binary, every feature (and
similarly, every sample) has a score between -1 and 1; the further the
score is from zero the stronger the association between the feature
(or sample) and the module. If two features both have scores with the
same sign, then they are correlated, if the sign of their scores are
opposite, then they are anti-correlated. You can query the scores of
the features with the \Rfunction{getFeatureScores} function, and
similarly, the \Rfunction{getSampleScores} function queries the sample
scores. You can supply the modules you want to query as an optional
argument:
<<query scores>>=
getFeatureScores(modules, 3)
getSampleScores(modules, 3)
@

@@

Objects from the \Rclass{ISAModules} class store various information
about the ISA run and the convergence of the seeds. Information
associated with the individual seeds can be queried with the
\Rfunction{seedData} function, it returns a data frame, with as many
rows as the number of seeds and various seed-level information,
e.g. the number of iterations required for the seed to converge. See
the manual page of \Rfunction{isa} for details.
<<seed data>>=
seedData(modules)
@

@@

The \Rpackage{eisa} package provides some functions to perform
enrichment tests for the genes in the various modules against various
databases. These tests are usually simplified and less sophisticated
versions than the ones in the \Rpackage{Category}, \Rpackage{GOstats}
or \Rpackage{topGO} packages, but they are much faster and this is
important if we need to perform them for many modules.

@@

To perform enrichment analysis against the Gene Ontology database, all
you have to do is to supply your \Rclass{ISAModules} object to the
\Rfunction{ISA.GO} function.
<<GO enrichment>>=
GO <- ISA.GO(modules)
@

The \Rfunction{ISA.GO} requires an annotation package that maps the
Entrez identifiers of the studied organism to GO terms. For human, the
\Rpackage{org.Hs.eg.db} package is required, for the mouse the
\Rpackage{org.Mm.eg.db} package, etc.

The \variable{GO} object is a list with three elements, these
correspond to the ontologies: biological function, cellular component
and molecular function, in this order.
<<list GO result>>=
GO
@

@@

For extracting the results themselves, the \Rfunction{summary}
function can be used, this converts them to a simple data frame. The
$p$-value limit can be supplied to \Rfunction{summary}. Note, that
since \Rfunction{ISA.GO} calculates enrichment for many gene sets
(i.e. for all biclusters), \Rfunction{summary} returns a list of data
frames, one for each bicluster. A table for the first module:
<<GO summary>>=
summary(GO[[1]], p=0.001)[[2]][,-6]
@

We omitted the sixth column of the result, because it is very wide and
would look bad in this vignette. This column is called
\variable{drive} and lists the Entrez IDs of the genes that are in the
intersection of the bicluster and the GO category; or in other
words, the genes that drive the enrichment.
These genes can also be obtained with the
\Rfunction{geneIdsByCategory} function. The following returns the
genes in the second module and the third GO BP category. (The GO
categories are ordered according to the enrichment $p$-values, just
like in the output of \Rfunction{summary}.)
<<GO gene ids by category>>=
geneIdsByCategory(GO[[1]])[[2]][[3]]
@

You can use the \Rpackage{GO.db} package to obtain more information
about the enriched GO categories.
<<GO info>>=
sigCategories(GO[[1]])[[2]]
library(GO.db)
mget(sigCategories(GO[[1]])[[2]], GOTERM)
@

In addition, the following functions are implemented to work on the
objects returned by \Rfunction{ISA.GO}: \Rfunction{htmlReport},
\Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
\Rfunction{exprectedCounts}, \Rfunction{universeCounts},
\Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
\Rfunction{geneIdUniverse}. These functions do essentially the same as
they counterparts for \Rclass{GOHyperGResult} objects, see the
documentation of the \Rpackage{GOstats} package. The only difference
is, that since here we are testing a list of gene sets (=biclusters),
they calculate the results for all gene sets and return a list.

@@

By default, the \Rfunction{ISA.GO} function performs multiple testing
correction using the Holm method, this can be changed via the
\texttt{correction} and \texttt{correction.method} arguments. See the
manual page of the \Rfunction{p.adjust} function for the possible
multiple testing correction schemes.

@@

Enrichment calculation against the KEGG pathway goes essentially the
same way as for the Gene Ontology, this time we use the
\Rfunction{ISA.KEGG} function:
<<KEGG enrichment>>=
KEGG <- ISA.KEGG(modules)
KEGG
summary(KEGG)[[4]]
library(KEGG.db)
mget(sigCategories(KEGG)[[4]], KEGGPATHID2NAME)
@

@@

The \Rpackage{eisa} includes a simple way to check whether the genes
in a bicluster are associated with a chromosome. See the
\Rfunction{ISA.CHR} function:
<<CHR enrichment>>=
CHR <- ISA.CHR(modules)
summary(CHR,p=0.05)[[2]][,-6]
@

@@

$\mu$RNAs are short RNA molecules that regulate gene expression.
TargetScan is a data based of predicted target genes of $\mu$RNAs, for
several organisms. There are two R packages that incorporate this
database, one for human and another one for mouse, right now they can
be downloaded from
\url{http://www2.unil.ch/cbg/index.php?title=Building_BioConductor_Annotation_Packages}.
The \Rpackage{targetscan.Hs.eg.db} package is for human, the
\Rpackage{targetscan.Mm.eg.db} package is for mouse.

The enrichment calculation itself is basically the same as for GO and
KEGG, but the \Rfunction{ISA.miRNA} function should be used:
<<miRNA enrichment>>=
if (require(targetscan.Hs.eg.db)) {
  miRNA <- ISA.miRNA(modules)
  summary(miRNA, p=0.1)[[7]]
}
@

@@

\subsection{The \Rpackage{biclust} package}

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{ISA.biclust} function converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
Bc <- ISA.biclust(modules)
Bc
@

@@

The easiest way to create a heatmap of a single module is to call the
\Rfunction{ISA2heatmap} function. You need to specify which module you
want to plot and also the \Rclass{ExpressionSet} object that is being
analyzed. Note that by default ISA normalizes the expression data
before running the module detection; the raw, non-normalized values
are plotted in the next example. If you want to plot the normalized
values, then you need to do the normalization ``by hand'', before
calling ISA, see Section~\ref{sec:detailed-isa}.
<<heatmap,fig=TRUE>>=
ISA2heatmap(modules, 1, ALL)
@

@@

You can also use the \Rpackage{biclust} package to create image
plots, by calling the \Rfunction{drawHeatmap} function. The result is
in Fig.~\ref{fig:drawHeatmap}.
<<heatmap biclust pre,echo=FALSE,results=hide>>=
if (require(Cairo)) {
  CairoPNG("graphics/heatmap.png", 400, 400 )  
} else {
  png("graphics/heatmap.png", 400, 400 )
}
@ 
<<heatmap>>=
drawHeatmap(exprs(ALL[featureNames(modules),]), Bc, number=1) 
@
<<heatmap post,echo=FALSE,results=hide>>=
dev.off() 
@ 
\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{graphics/heatmap}
  \caption{Heatmap of the first transcription module.}
  \label{fig:drawHeatmap}
\end{figure}

@@

Profile plots visualize the difference between the genes (or samples)
in the modules and the rest of the expression data. A profile plot
contains a line plot for every single gene (or sample) and the genes
that belong to the module have a different color, see
Fig.~\ref{fig:parallel}.
<<parallel.pre,echo=FALSE,results=hide>>=
if (require(Cairo)) {
  CairoPNG("graphics/parallel.png", 400, 400 )
} else {
  png("graphics/parallel.png", 400, 400 )
}
@ 
<<parallel>>=
profile.plot(modules, 2, ALL, plot="both")
@
<<parallel.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{graphics/parallel}
\caption{Profile plots for the second bicluster found by ISA.}
\label{fig:parallel}
\end{figure}

@@

The \Rpackage{eisa} package provides some functions to plot parts of
the GO graph that is related to a transcription module. The
\Rfunction{gograph} function creates an object that is a
representation of such a plot. Its input is a table with the GO
categories to plot and their enrichment $p$-values. (Additional
columns are silently ignored.) Here is how to use
it on the previously calculated enrichment scores:
<<goplot1>>=
goplot.2 <- gograph(summary(GO[[1]], p=0.05)[[1]])
@

\Rfunction{goplot} uses the \Rpackage{igraph} package to create a
graph with associated meta data:
<<gograph object>>=
summary(goplot.2)
list.vertex.attributes(goplot.2)
@ 
The \texttt{width} and \texttt{height} graph attributes contain the
suggested width and height of the graph, if plotted to a bitmap
device. (The graph attributes of an \Rpackage{igraph} graph can be
queried with the `\texttt{\$}' selector:
<<gograph object 2>>=
goplot.2$width
goplot.2$height
@

Let's plot the graph, we can do this with the \Rfunction{gograph.plot}
function.
<<goplot2,eval=FALSE>>=
x11(width=10, height=10*goplot.2$height/goplot.2$width)
gograph.plot(goplot.2)
@
\setkeys{Gin}{width=\textwidth}
<<goplot2-real,echo=FALSE,width=10,height=2.5,fig=TRUE>>=
ps.options(fonts=c("serif", "mono"))
gograph.plot(goplot.2)
@

Because the GO is not really a tree, \Rfunction{gograph} ``unfolds''
it into a tree by including categories more than once, if needed. It
also abbreviates the names of the GO categories to make them fit on
the tree. The graph object contains the full name of the category as
well. The full and abbreviated names can be listed by querying the
appropriate vertex attributes of the graph. Here they are for the
first five categories:
<<goplot names>>=
V(goplot.2)$abbrv[1:5]
lapply(V(goplot.2)$desc[1:5], strwrap)
@ 
\Rfunction{gograph.plot} colors the categories according to the
supplied enrichment $p$-values, the minus $\log_{10}$ $p$-value is
also added to the plot, see the bold blue numbers.

@@

The \Rfunction{cond.plot} function can plot the sample scores,
potentially including the scores before the ISA filtering. Let us plot
the scores for the second module, the color code denotes B-cell
vs. T-cell leukemia.

<<condplot,fig=TRUE,width=10,height=4>>=
col <- ifelse( grepl("^B", pData(modules)$BT), "orange", "darkolivegreen")
cond.plot(modules, 2, ALL, col=col)
@

@@

The \Rfunction{autogen.table} function creates a summary

\section{How ISA works}

Before showing an actual ISA tool chain, a few words about how the
algorithm works are in order.

@@

It is very possible that the ISA finds the same module more than once;
two or more seeds might converge to the same module. The function
\Rfunction{ISA.unique} eliminates every module from the result of 
\Rfunction{ISA.iterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

@@

By default (i.e. if the \Rfunction{isa} function is used) the ISA is
performed from random sparse starting seeds, generated by the
\Rfunction{generate.seeds} function. This way the algorithm is 
completely unsupervised, but also stochastic: it might give different
results for different runs.

It is possible to use non-random seeds as well. If you have some
knowledge about the data or are interested in a particular subset of
genes/samples, then you can feed in your seeds into the
\Rfunction{ISA.iterate} function directly. In this case the
algorithm is deterministic, for the same seed you will always get the
same results. Using smart (i.e. non-random) seeds can be considered as
a semi-supervised approach.

@@

@@

>>>>

\newcommand{\Rfunction}[1]{\texttt{#1()}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\variable}[1]{\texttt{#1}}

\SweaveOpts{cache=TRUE}
\SweaveOpts{prefix.string=plot}

\setlength{\parindent}{2em}

@@

<<set width,echo=FALSE,print=FALSE>>=
options(width=60)
options(continue=" ")
@

@@

The Iterative Signature Algorithm (ISA) is biclustering method. The
input of a biclustering method a matrix and the output is a set of
biclusters that fulfill some criteria. A bicluster is a block of the
potentially reordered input matrix.

Most commonly, this algorithm is used on microarray expression data,
to find gene sets that are coexpressed across a subset of the original
samples. In the ISA papers the biclusters are called
transription modules (TM), we will often refer them under this name in
the following.

This tutorial specifically deals with the modular analysis of gene
expression data. Section~\ref{sec:isa} gives a short summary of how
ISA works. If you need more information of the underlying math
or want to apply it to other data, then please see the referenced
papers, the vignette titled ``The Iterative Signature Algorithm'' in
the \Rpackage{isa2} R package, or the ISA homepage at
\url{http://www.unil.ch/cbg/homepage/software.html}.

@@

First, we load the required packages and the data to analyze. ISA is
implemented in the \Rpackage{eisa} and \Rpackage{isa2} packages, see
Section~\ref{sec:isapackages} for a more elaborated summary about the
two packages. It is enough to load the \Rpackage{eisa} package,
\Rpackage{isa2} and other required packages are loaded automatically:
<<load the packages,cache=FALSE>>=
library(eisa)
@

In this tutorial we will use the data in the \Rpackage{ALL}
package. 
<<load the data,cache=FALSE>>=
library(ALL)
library(hgu95av2.db)
library(affy)
data(ALL)
@
This is a data set from a clinical trial in acute
lymphoblastic leukemia and it contains \Sexpr{ncol(ALL)} samples
altogether.

@@

The simplest way to run ISA is to choose the two threshold parameters
and then call the \Rfunction{isa} function on the
\Rclass{ExpressionSet} object. The threshold parameters tune the size
of the modules, less stringent (i.e. smaller) values result bigger,
less correlated modules. The optimal values depend on your data and
some experimentation is needed to determine them.

Since running ISA might take a couple of minutes and the results 
depend on the random number generator used, the ISA run is commented out 
and we just load a precomputed set of modules in this vignette.
<<simple ISA>>=
thr.gene <- 2.7
thr.cond <- 1.4
set.seed(1) # to get the same results, always
# modules <- ISA(ALL, thr.gene=thr.gene, thr.cond=thr.cond)
data(ALLModulesSmall)
modules <- ALLModulesSmall
@

This first applies a non-specific filter to the data set and then runs
ISA from \Sexpr{formals(eisa:::ISA)$no.seeds} % $
random seeds (the default). See Section~\ref{sec:detailed-isa} if the
default parameters are not appropriate for you and need more control.

@@

Calling \Rfunction{length} on \variable{modules} returns the number of
ISA modules in the set, \Rfunction{dim} gives the dimension of the
input expression matrix: the number of features (after the filtering)
and the number of samples:
<<accessors 1>>=
length(modules)
dim(modules)
@

Functions \Rfunction{featureNames} and \Rfunction{sampleNames} return
the names of the features and samples, just like the functions with
the same name for an \Rclass{ExpressionSet}:
<<accessors 2>>=
featureNames(modules)[1:5]
sampleNames(modules)[1:5]
@

The \Rfunction{getNoFeatures} function returns a numeric vector, the
number of features (probesets in our case) in each module. Similarly,
\Rfunction{getNoSamples} returns a numeric vector, the number of
samples in each module. \Rfunction{pData} returns the
phenotype data of the expression set as a data frame. The
\Rfunction{getOrganism} function returns the scientific name of the
organism, \Rfunction{annotation} the name of the chip. For the former
the appropriate annotation package must be installed.
<<number of features and samples>>=
getNoFeatures(modules)
getNoSamples(modules)
colnames(pData(modules))
getOrganism(modules)
annotation(modules)
@

@@

\Rfunction{getFeatureNames} lists the probesets (more precisely,
the feature names coming from the \Rclass{ExpressionSet} object) in
the modules. It returns a list, here we just print the first entry in
the list.
<<list genes in modules>>=
getFeatureNames(modules)[[1]]
@

@@

ISA biclustering is not binary, every feature (and
similarly, every sample) has a score between -1 and 1; the further the
score is from zero the stronger the association between the feature
(or sample) and the module. If two features both have scores with the
same sign, then they are correlated, if the sign of their scores are
opposite, then they are anti-correlated. You can query the scores of
the features with the \Rfunction{getFeatureScores} function, and
similarly, the \Rfunction{getSampleScores} function queries the sample
scores. You can supply the modules you want to query as an optional
argument:
<<query scores>>=
getFeatureScores(modules, 3)
getSampleScores(modules, 3)
@

@@

Objects from the \Rclass{ISAModules} class store various information
about the ISA run and the convergence of the seeds. Information
associated with the individual seeds can be queried with the
\Rfunction{seedData} function, it returns a data frame, with as many
rows as the number of seeds and various seed-level information,
e.g. the number of iterations required for the seed to converge. See
the manual page of \Rfunction{ISA} for details.
<<seed data>>=
seedData(modules)
@

@@

The \Rpackage{eisa} package provides some functions to perform
enrichment tests for the gene sets corresponding to the ISA modules
against various databases. These tests are usually simplified and less
sophisticated versions than the ones in the \Rpackage{Category},
\Rpackage{GOstats} or \Rpackage{topGO} packages, but they are much
faster and this is important if we need to perform them for many
modules.

@@

To perform enrichment analysis against the Gene Ontology database, all
you have to do is to supply your \Rclass{ISAModules} object to the
\Rfunction{ISAGO} function.
<<GO enrichment>>=
GO <- ISAGO(modules)
@

The \Rfunction{ISAGO} function requires the annotation package on the
chip, e.g. for the ALL data, the \Rpackage{hgu95av2.db} package is
required.

The \variable{GO} object is a list with three elements, these
correspond to the GO ontologies, they are: biological function,
cellular component and molecular function, in this order.
<<list GO result>>=
GO
@

@@

For extracting the results themselves, the \Rfunction{summary}
function can be used, this converts them to a simple data frame. A
$p$-value limit can be supplied to \Rfunction{summary}. Note, that
since \Rfunction{ISAGO} calculates enrichment for many gene sets
(i.e. for all biclusters), \Rfunction{summary} returns a list of data
frames, one for each bicluster. A table for the first module:
<<GO summary>>=
summary(GO$BP, p=0.001)[[1]][,-6]
@

We omitted the sixth column of the result, because it is very wide and
would look bad in this vignette. This column is called
\variable{drive} and lists the Entrez IDs of the genes that are in the
intersection of the bicluster and the GO category; or in other
words, the genes that drive the enrichment.
These genes can also be obtained with the
\Rfunction{geneIdsByCategory} function. The following returns the
genes in the first module and the third GO BP category. (The GO
categories are ordered according to the enrichment $p$-values, just
like in the output of \Rfunction{summary}.)
<<GO gene ids by category>>=
geneIdsByCategory(GO$BP)[[1]][[3]]
@

You can use the \Rpackage{GO.db} package to obtain more information
about the enriched GO categories.
<<GO info,cache=FALSE>>=
sigCategories(GO$BP)[[1]]
library(GO.db)
mget(sigCategories(GO$BP)[[1]], GOTERM)
@

In addition, the following functions are implemented to work on the
objects returned by \Rfunction{ISAGO}: \Rfunction{htmlReport},
\Rfunction{pvalues}, \Rfunction{geneCounts}, \Rfunction{oddsRatios},
\Rfunction{exprectedCounts}, \Rfunction{universeCounts},
\Rfunction{universeMappedCount}, \Rfunction{geneMappedCount},
\Rfunction{geneIdUniverse}. These functions do essentially the same as
they counterparts for \Rclass{GOHyperGResult} objects, see the
documentation of the \Rpackage{GOstats} package. The only difference
is, that since here we are testing a list of gene sets (=biclusters),
they calculate the results for all gene sets and usually return lists.

@@

By default, the \Rfunction{ISAGO} function performs multiple testing
correction using the Holm method, this can be changed via the
\texttt{correction} and \texttt{correction.method} arguments. See the
manual page of the \Rfunction{ISAGO} function for details, and also
the \Rfunction{p.adjust} function for the possible multiple testing
correction schemes.

@@

Enrichment calculation against the KEGG pathway database goes
essentially the same way as for the Gene Ontology, this time we use
the \Rfunction{ISAKEGG} function:
<<KEGG enrichment,cache=FALSE>>=
KEGG <- ISAKEGG(modules)
KEGG
summary(KEGG, p=1e-3)[[1]]
library(KEGG.db)
mget(sigCategories(KEGG)[[1]], KEGGPATHID2NAME)
@

@@

The \Rpackage{eisa} includes a simple way to check whether the genes
in a bicluster are associated with a chromosome. See the
\Rfunction{ISACHR} function:
<<CHR enrichment>>=
CHR <- ISACHR(modules)
summary(CHR,p=0.05)[[2]][,-6]
@

@@

$\mu$RNAs are short RNA molecules that regulate gene expression.
TargetScan is a database of predicted target genes of $\mu$RNAs, for
several organisms. There are two R packages that incorporate this
data, one for human and another one for mouse, right now they can
be downloaded from
\url{http://www.unil.ch/cbg/homepage/software.html}.
The \Rpackage{targetscan.Hs.eg.db} package is for human, the
\Rpackage{targetscan.Mm.eg.db} package is for mouse.

The enrichment calculation itself is basically the same as for GO and
KEGG, but the \Rfunction{ISAmiRNA} function should be used:
<<miRNA enrichment>>=
if (require(targetscan.Hs.eg.db)) {
  miRNA <- ISAmiRNA(modules)
  summary(miRNA, p=0.1)[[7]]
}
@

@@

\subsection{The \Rpackage{biclust} package}%
\label{sec:biclust}

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The standard R \Rfunction{as} function
can be used to convert ISA modules to a \Rclass{Biclust} object. This
requires the binarization of the modules, i.e. the ISA scores are
lost, they are converted to zeros and ones:
<<biclust,cache=FALSE>>=
library(biclust)
Bc <- as(modules, "Biclust")
Bc
@

@@

The easiest way to create a heatmap of a single module is to call the
\Rfunction{ISA2heatmap} function. You need to specify which module you
want to plot and also the \Rclass{ExpressionSet} object that is being
analyzed. Note that by default ISA normalizes the expression data
before running the module detection, but \Rfunction{ISA2heatmap} plots
the raw values by default, optionally scaled. 
If you want to plot the normalized values, then you need supply the
\Rargument{norm} argument to \Rfunction{ISA2heatmap}, see the manual
for details. The result is in Fig.~\ref{fig:heatmap}.
<<heatmap,results=hide,include=FALSE,eval=FALSE>>=
ISA2heatmap(modules, 1, ALL, norm="sample", scale="none")
@

\begin{figure}
\centering
<<heatmap-real,fig=TRUE,echo=FALSE,results=hide>>=
<<heatmap>>
@
\caption{Heatmap of the first ISA module, plotted via the
\Rfunction{ISA2heatmap} function. The expression values are
sample-wised normalized.}
\label{fig:heatmap}
\end{figure}

@@

You can also use the \Rpackage{biclust} package to create image
plots, by calling the \Rfunction{drawHeatmap} function. The result is
in Fig.~\ref{fig:drawHeatmap}.
<<heatmap biclust pre,echo=FALSE,results=hide>>=
if (require(Cairo)) {
  CairoPNG("plot-heatmap.png", 400, 400 )  
} else {
  png("plot-heatmap.png", 400, 400 )
}
@ 
<<heatmap>>=
drawHeatmap(exprs(ALL[featureNames(modules),]), Bc, number=1) 
@
<<heatmap post,echo=FALSE,results=hide>>=
dev.off() 
@ 
\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{plot-heatmap.png}
  \caption{Heatmap of the first transcription module, plotted with the
  \Rfunction{drawHeatmap} function from the \Rpackage{biclust}
  package.}
  \label{fig:drawHeatmap}
\end{figure}

@@

Profile plots visualize the difference between the genes (or samples)
in the modules and the rest of the expression data. A profile plot
contains a line plot for every single gene (or sample) and the genes
that belong to the module have a different color, see
Fig.~\ref{fig:parallel}.
<<parallel.pre,echo=FALSE,results=hide>>=
if (require(Cairo)) {
  CairoPNG("plot-parallel.png", 400, 400 )
} else {
  png("plot-parallel.png", 400, 400 )
}
@ 
<<parallel>>=
profilePlot(modules, 2, ALL, plot="both")
@
<<parallel.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{plot-parallel.png}
\caption{Profile plots for the second bicluster found by ISA.}
\label{fig:parallel}
\end{figure}

@@

The \Rpackage{eisa} package provides functions to plot parts of
the GO graph that is related to a transcription module. The
\Rfunction{gograph} function creates an object that is a
representation of such a plot. Its input is a table with the GO
categories to plot and their enrichment $p$-values. (Additional
columns are silently ignored.) Here is how to use
it on the previously calculated enrichment scores:
<<goplot1>>=
goplot.2 <- gograph(summary(GO$BP, p=0.05)[[1]])
@ 
% $

\Rfunction{goplot} uses the \Rpackage{igraph} package to create a
graph with the associated meta data:
<<gograph object>>=
summary(goplot.2)
@ 
The \texttt{width} and \texttt{height} graph attributes contain the
suggested width and height of the graph in pixels, if plotted to a bitmap
device. (The graph attributes of an \Rpackage{igraph} graph can be
queried with the `\texttt{\$}' selector:
<<gograph object 2>>=
goplot.2$width
goplot.2$height
@

Let's plot the graph, we can do this with the \Rfunction{gographPlot}
function, see Fig.~\ref{fig:gotree}.
<<goplot2,eval=FALSE>>=
x11(width=10, height=10*goplot.2$height/goplot.2$width)
gographPlot(goplot.2)
@
\setkeys{Gin}{width=\textwidth}
<<goplot2-real,results=hide,include=FALSE,eval=FALSE,echo=FALSE>>=
ps.options(fonts=c("serif", "mono"))
gographPlot(goplot.2)
@

\begin{figure}
\centering
<<goplot2-real-real,width=10,height=5,fig=TRUE,results=hide,echo=FALSE,cache=FALSE>>=
<<goplot2-real>>
@
\caption{Part of the Gene ontology database, ``Biological Function''
ontology, that contains all GO terms enriched by a transcription module.}
\label{fig:gotree}
\end{figure}

Because the GO is not a tree, \Rfunction{gograph} ``unfolds''
it into a tree by including categories more than once, if needed. It
also abbreviates the names of the GO categories to make them fit on
the plot. The graph object contains the full name of the category as
well. The full and abbreviated names can be listed by querying the
appropriate vertex attributes of the graph. Here they are for the
first five categories:
<<goplot names>>=
V(goplot.2)$abbrv[1:5]
lapply(V(goplot.2)$desc[1:5], strwrap)
@ 
\Rfunction{gographPlot} colors the categories according to the
supplied enrichment $p$-values, the minus $\log_{10}$ $p$-value is
also added to the plot, see the bold blue numbers.

@@

The \Rfunction{condPlot} function can plot the sample scores,
potentially including the scores before the ISA filtering. Let us plot
the scores for the second module, the color code denotes B-cell
vs. T-cell leukemia (Fig.~\ref{fig:condplot}).

<<condplot,results=hide,include=FALSE,eval=FALSE>>=
col <- ifelse( grepl("^B", pData(modules)$BT), "darkolivegreen", "orange")
condPlot(modules, 2, ALL, col=col)
@
% $

\begin{figure}
\centering
<<condplot-real,fig=TRUE,width=10,height=4,echo=FALSE,results=hide>>=
<<condplot>>
@
\caption{Condition plot for the ALL data and the second module. Every
  sample is represented as a bar, and its ISA score is plotted. The
  scores are also printed at the top or bottom of the bars. The scores
  corresponding to the ISA thresholds are represented as horizontal
  lines. The dashed line is the mean of the condition scores. B-cell
  samples are dark green, T-cell samples are orange.
}
\label{fig:condplot}
\end{figure}

@@

The \Rfunction{ISAHTMLTable} function creates a summary of a set of
ISA modules, as an HTML table. The summary optionally includes the
results of the enrichment calculations as well. Fig.~\ref{fig:HTML}
shows a screenshot of the table created for the
\Sexpr{length(modules)} modules we have found earlier.
<<HTML1>>=
htmldir <- tempdir()
ISAHTMLTable(modules=modules, target.dir=htmldir, 
             GO=GO, KEGG=KEGG, CHR=CHR)
@ 
<<HTML2,eval=FALSE>>=
if (interactive()) {
  browseURL(URLencode(paste("file://", htmldir, "/maintable.html", sep="")))
}
@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{plot-HTMLTable.png}
\caption{A HTML summary table showing information for the first couple
of modules found in the ALL data set. The ``Thr.'' column shows the
ISA thresholds that were used to find the module; the next two columns
are the number of features (genes) and the number of samples
(conditions) in the modules. The other columns show the most
significantly enriched GO categories, KEGG pathways and
chromosomes. The small grey numbers are the enrichment $p$-value, the
number of genes expected to be hit by chance, the number of hits and
the size of the category (within the respective gene universe), in
this order.}
\label{fig:HTML}
\end{figure}

The Rfunction{ISAHTMLModules} generates a HTML page for every single
module with the following information on it: 
\begin{itemize}
\item An expression plot of the genes and samples in the module,
  including the ISA scores. This is done by calling \Rfunction{expPlot}.
\item Overlap plot of all modules, see \Rfunction{overlapPlot}.
\item  Gene Ontology tree plots for the enriched GO terms,
  separately for the three ontologies., These are produced by
  calling \Rfunction{gograph}.
\item Tables for the enriched Gene Ontology terms, separately for
  the three ontologies.
\item A table for the enriched KEGG pathways.
\item A table for the enriched miRNA families.
\item The list of genes in the module.
\item The list of samples in the module.
\item A condition plot, see the \Rfunction{condPlot} function.
\end{itemize}
<<HTML modules>>=
ISAHTMLModules(eset=ALL, modules=modules, target.dir=htmldir,
               GO=GO, KEGG=KEGG, CHR=CHR)
@ 
The rows of the summary table generated by \Rfunction{ISAHTMLTable}
link to the module pages generated by \Rfunction{ISAHTMLModules}.

Both the summary and the module pages can be generated with the
\Rfunction{ISAHTML} function.

\subsection{The ExpressionView package}

The previously mentioned visualization techniques focus on a single
module only. The \Rpackage{ExpressionView} package visualizes a set of
possibly overlapping biclusters together.

The \Rpackage{ExpressionView} package~\cite{ev} tries to reorder the
rows and columns of the expression matrix in such a way that the
genes/samples that appear together in a module, are placed right
beside/above each other. The package directly supports the
\Rclass{ISAModules} and \Rclass{ExpressionSet} classes, see its
documentation for details.

\section{How ISA works}%
\label{sec:isa}

@@

It is very possible that the ISA finds the same module more than once;
two or more seeds might converge to the same module. The function
\Rfunction{ISAUnique} eliminates every module from the result of 
\Rfunction{ISAIterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

@@

By default (i.e. if the \Rfunction{ISA} function is used) the ISA is
performed from random sparse starting seeds, generated by the
\Rfunction{generate.seeds} function. This way the algorithm is 
completely unsupervised, but also stochastic: it might give different
results for different runs.

It is possible to use non-random seeds as well. If you have some
knowledge about the data or are interested in a particular subset of
genes/samples, then you can feed in your seeds into the
\Rfunction{ISAIterate} function directly. In this case the
algorithm is deterministic, for the same seed you will always get the
same results. Using smart (i.e. non-random) seeds can be considered as
a semi-supervised approach. We show an example of using smart seeds in
Section~\ref{sec:detailed-isa}.

@@

As it turns out, one modules found using the smart seeds was also
found with random seeds.

@@

\section{More information}

For more information about the ISA, please see the references
below. The ISA homepage at
\url{http://www.unil.ch/cbg/homepage/software.html} has example data
sets, and all ISA related tutorials and papers.
